---
title: "LAB 1"
subtitle: "Applied Survival Analysis"
author: "Sarah C. Lotspeich, Ph.D."
date: "Spring 2023"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival) #datasets and functions
library(dplyr)
library(ggplot2)
library(tableone)
library(magrittr)
```

### Preamble: To Pipe or Not to Pipe

```{r}
# Take the square root of the sequence 1-10 and then log transform

## Without pipes
log(sqrt(1:10))

## With pipes
1:10 %>%
  sqrt() %>%
  log()
```


### Introduction to the `veteran` data

The `survival` package contains data from Veterans’ Administration Lung Cancer study (i.e., the `veteran` dataset). This was a randomized trial of two treatment regimens for lung cancer, and `r nrow(veteran)` patients participated. Recruited sometime after diagnosis, patients were followed until death or the end of the study. Additional information such as age and tumor type were also collected. 

A brief data dictionary is given in the documentation:

  -   `trt`: treatment regimen (`1` = standard, `2` = test)
  -   `celltype`: tumor type (`squamous`, `smallcell`, `adeno`, `large`)
  -   `time`: time from treatment to either death or censoring (in days)
  -   `status`: event status (`0` = censored, `1` = dead)
  -   `karno`: Karnofsky performance score (100=good)
  -   `diagtime`: time from diagnosis to treatment (in months)
  -   `age`: in years
  -   `prior`: prior therapy (`0` = no, `10` = yes)

These data were sourced from Kalbfleisch \& Prentice (1980), \textit{The Statistical Analysis of Failure Time Data}. Wiley, New York.

Recall that in survival analysis, our data need to have three characteristics:
(1) a clear time origin, (2) a defined time scale, and (3) an event of interest. 
  
#### *Your turn:* Identify these characteristics in the `veteran` data. 

  1.    Time origin: treatment
  2.    Time scale: days
  3.    Event of interest: death or end of study

#### Import the data

Once you've loaded the `survival` package, you already have the `veteran` data! Don't believe me? Try to use it. 

```{r}
head(veteran)
```

#### Add labels for `trt` and `prior`

```{r label}
# Turn trt and prior into labeled factor variables
veteran <- veteran %>% mutate(trt = as.factor(trt), prior = as.factor(prior))
glimpse(veteran)
```

Note: `status` is also categorical, but we want to leave it as an integer for some survival-specific functions to come. 

### Exploratory data analysis

#### Staggered entry

In our first lecture, we mentioned [**staggered entry**](https://dictionary.apa.org/staggered-entry-design) as one of the common features in survival studies. In the `veteran` dataset, patients weren't recruited into the study at the exact time of their diagnosis but some time after. Fortunately, this information was captured in the `diagtime` variable. 

Create either a graphical or numerical summary of how many patients entered the study when (relative to their diagnoses). 

```{r}
# Summarize who entered the study when 
when <- mean(veteran$diagtime)
veteran %>% ggplot(aes(x = diagtime)) + 
  geom_histogram(fill = "blue", bins = 30) +
  geom_vline(xintercept = when, color = "red") +
  theme_bw() +
  labs(title = "Distribution of Entry Times",
       subtitle = "Mean = 8.77 months",
       x = "Time from Diagnosis to Treatment",
       y = "Count")

```

#### Patient demographics 

Clinical papers often include a "Table 1," which describes the demographics of the patients in the study. You can do this manually, but for fun I thought I'd also introduce you to the `tableOne` package, which does this super nicely. 

You can create an overall summary of all patients, 

```{r}
# Create table summarizing patients (overall)
tableOne <- CreateTableOne(data = veteran)
tableOne
```

or broken down (i.e., stratified) on something like treatment group

```{r}
# Create table summarizing patients (stratified by trt)
tableOneStrat <- CreateTableOne(data = veteran, strata = c("trt"))
tableOneStrat
```

#### Event frequency 

```{r}
# Tabulate event frequency
veteran$stat[which(veteran$status == "0")] <- "Censored"
veteran$stat[which(veteran$status == "1")] <- "Dead"
veteran %>% mutate(stat = as.factor(stat)) %>% select(stat) %>% table()
```

Of `r nrow(veteran)` patients in the trial, `r sum(veteran$status)` (`r round(mean(veteran$status) * 100)`%) died before the study ended, leaving just `r nrow(veteran) - sum(veteran$status)` (`r round(mean(veteran$status == 0) * 100)`%) censored. We can look at the event frequency further broken down by other risk factors, like... 

```{r}
# Tabulate event frequency by treatment group
veteran %>% select(stat, trt) %>% table()
```

  -   **By treatment group:** There were `r with(veteran, table(status, trt))[2, 1]` deaths among the standard treatment group (`r with(veteran, table(status, trt))[2, 1]/sum(with(veteran, table(status, trt))[, 1])`) and `r with(veteran, table(status, trt))[2, 2]/sum(with(veteran, table(status, trt))[, 2])` among the test treatment group.

```{r}
# Tabulate event frequency by prior therapy status
veteran %>% select(stat, prior) %>% table()
```

  -   **By prior therapy status:** There were `r with(veteran, table(status, prior))[2, 1]` deaths among people without prior therapy (`r with(veteran, table(status, prior))[2, 1]/sum(with(veteran, table(status, prior))[, 1])`) and `r with(veteran, table(status, prior))[2, 2]/sum(with(veteran, table(status, prior))[, 2])` among people with prior therapy.

We could also do this for `celltype`, but there are four levels so interpretations would be a little more cumbersome.

### Preparing for estimation
#### Create time-to-event variable

We got lucky here - the time-to-event variable (`time`) was already calculated for us. Other times we might instead have `date`-type variables and would have to calculate the time-to-event variable from them.

#### Create survival object

Finally we create a special type of R object called a `Surv`, which contains the two types of information needed for survival data: the time-to-event and the event indicator.

```{r surv_obj}
# Create Surv object for the veteran data
# Create Surv object for the veteran data
S = Surv(time = veteran$time, 
          event = veteran$status)
S
```


The `Surv` object is a key input to many of the `survival` package functions for estimation and modeling. For example, we could feed `S` into the `survfit()` function to obtain a ``nonparametric'' (i.e., data-driven, distribution-free) estimate of the survival function. 

```{r}
# Example: Fit a nonparametric survival curve for the veteran data
## You can feed S in from above, in which case you don't need to specify the "data" argument
KM <- survfit(formula = S ~ 1)
KM 

## Or you can define the Surv object inside of the survfit() call
KM <- survfit(formula = Surv(time = time, event = status) ~ 1, 
             data = veteran)


# Examine estimated survival with plot 
plot(KM) 
```

### Built-in functions for common distributions

Suppose that, instead of the nonparametric estimate, we have prior knowledge that the `time` to death in the `veteran` data set follows a **log-normal distribution with mean $\mu = 4.2$ and standard deviation $\sigma = 1.4$ (on the log-scale)**. This means that we can use many of the built-in functions for this distribution that come with ``base R.'' 

#### Quantity 1: the PDF

The PDF $f(x)$ is already implemented as the `dlnorm()` function, where the arguments are:

```{r}
# Compute the PDF of time at 14 days
dlnorm(x = 14, meanlog = 4.2, sdlog = 1.4)

```

Note: If we instead had discrete time data, then the distribution function leading with `d` will be the PMF instead. 

#### Quantity 2: the Survival

And `plnorm()` is the CDF $F(x)$. Use it to write your own function for $S(x)$ below, and then compute the survival probability 1 year post-treatment. 

```{r}
# Define a function for the survival 
Sx = function(x) {
  1 - plnorm(q = x, meanlog = 4.2, sdlog = 1.4)
}

# Compute survival one year post-treatment
Sx(365)
```

Now, plot the survival function for `time` for 2 years post-treatment. (Hint: You might want to save this plot because we're going to add layers to it.)

```{r}
# Plot the survival function for t = 0 to 2 * 365
plot_Sx <- veteran %>% ggplot(aes(x = time, y = Sx(time))) + 
  geom_line() +
  labs(title = "Survival Past Treatment",
       x = "Time Past Treatment",
       y = "Probability of Survival Past That Time") +
  theme_bw()

plot_Sx
```

How long post-treatment is the survival probability 50\%? We can find the exact value using the quantile function `qlnorm()`: 

```{r}
Sx50 <- qlnorm(0.5, meanlog = 4.2, sdlog = 1.4)
  
Sx50
```

Annotate this time point on your plot of the survival curve (maybe using dashed lines, shading, etc).

```{r}
plot_Sx + geom_vline(xintercept = 66.68633, color = "magenta")
```

#### Quantity 3: Hazard function 

By definition, we have that for continuous `time` the hazard function is $h(x) = f(x) / S(x)$. We have functions for $f(x)$ and $S(x)$, so let's put them together to create a function for $h(x)$.

```{r}
# Define a function for the hazard 
hx = function(x) {
  dlnorm(x = x, meanlog = 4.2, sdlog = 1.4) / Sx(x)
}
```

What is the shape of the hazard function for death in the `veteran` data? You can focus on the first 2 years post treatment. (Hint: You might want to save this plot because we're going to add layers to it.)

```{r}
# Plot the hazard function for t = 0 to 2 * 365
plot_hx <- veteran %>% ggplot(aes(x = time, y = hx(time))) + 
  geom_line() +
  labs(title = "Risk Past Treatment",
       x = "Time Past Treatment",
       y = "Probability of Dying In The Next Instant If You Survived This Far") +
  theme_bw()

plot_hx
```

Are there any changepoints in the hazard? Find the times associated with them using built-in numerical optimization functions (e.g., `nlm()` or `optim()`). I'd recommend something small and close to, but not exactly, zero as your start value (e.g., 1 day post-treatment). 

```{r}
# Use nlm() to find x to maximize ht
## (i.e., to minimize -ht)
neghx <- function(x) {
  -1 * dlnorm(x = x, meanlog = 4.2, sdlog = 1.4) / Sx(x)
}
nlm(f = neghx, p = 1)$estimate
```

#### Quantity 4: Cumulative hazard function

Now that we have a nice function for the hazard, we just need to integrate under it to get the cumulative hazard function $H(x)$. (Hint: The `integrate()` function in R would do a really nice job of this.) 

```{r}
# Define a function for the cumulative hazard 
Hx = function(x) {
  integrate(f = hx, lower = 0, upper = x)
}
```

Use your function to calculate the cumulative hazard of death 100 days post-treatment, 

```{r}
# Calculate cumulative hazard at x = 30
Hx(30)
```

and use shading to illustrate your plot of the hazard function to show where this comes from. 

```{r}

shadeH <-subset(veteran, time < 30)

plot_hx +
    stat_function(geom = "area", 
                fun = hx, 
                xlim = c(0, 30), 
                fill = "magenta", 
                alpha = 0.7
                )
```

#### Quantity 5: Mean residual life 

We now have all of the pieces necessary to define a function for the mean residual life function, defined $$mrl(x) = \textrm{E}(X - x|X > x) = \frac{\int_{x}^{\infty}S(t)dt}{S(x)}.$$ Note: the `integrate()` function can handle infinite upper bounds. 

```{r}
# Define function for mean residual life 
mrl = function(x) {
  integrate(f = Sx, lower = x, upper = Inf)$value / Sx(x)
}
```

Calculate the mean residual life after already surviving 10, 50, 100, and 500 days post-treatment. 

```{r}
# Calculate mean residual life at x = 50
times <- c(10,50,100,500)
for(i in times) {
  print(mrl(i))
}
```

#### Quantity 6: Mean life 

For log-normal `time`, we know that the mean should be $\exp(4.2 + 1/2 * 1.4 ^ 2) = 177.6826$ days. Let's use that knowledge to check our functions for mean life!

**Approach 1:** Using the definition of expected value from probability, we can use the PDF to calculate mean life. 

```{r}
integrate(f = function(x) x * dlnorm(x = x, meanlog = 4.2, sdlog = 1.4), 
          lower = 0,
          upper = Inf)
```

**Approach 2:** Using the definition of expected value from survival, we can use the survival function to calculate mean life. 

```{r}
integrate(f = function(x) Sx(x = x), 
          lower = 0,
          upper = Inf)
```

**Approach 3:** Use the mean residual life function, since mean life is just mean residual life after already surviving 0 days. 

```{r}
mrl(x = 0)
```

#### Quantity 7: Percentiles of life

We have two ways to find the $(100)p$th percentile of a time-to-event variable: (1) by inverting the CDF at $p$ or (2) by inverting the survival at $(1 - p)$. 

**Approach 1: Inverting the CDF.** Just like there are built-in functions for the CDF in R, there are functions for the inverse CDFs (i.e., the quantile functions). Try using `qlnorm()` to find the $25$th percentile (i.e., $p = 0.25$) of `time` to death in the `veteran` data set.

```{r}
# Find median life using qlnorm()
qlnorm(p = 0.25, meanlog = 4.2, sdlog = 1.4)
```

**Approach 2: Inverting the survival.** We can treat the constraint that $S(\phi_{p}) = 1 - p$ as an objective function and use a solver like `uniroot()` to find the value of $\phi_{p}$ to make it true. 

```{r}
# Define the constraint that survival at \phi = 1 - p
constr = function(phi, p) {
  Sx(phi) - (1 - p)
}


# Use uniroot() to find phi to satisfy it
uniroot(f = constr, interval = c(0,1000), p = 0.25)$root
```
