---
title: "Table1-ImputeTrue-WeibullX (The Commentary Edition)"
output: pdf_document
---

This notebook comments the living daylights out of the script used to produce Table 1 in the "It's Integral" paper, which really sticks it to the nemesis and does not deserve the hatred of the dreaded Reviewer #2. It censors a Weibull covariate X and imputes with the true survival function S(X|Z). 

The block below loads the appropriate packages and defines our experiment settings and ensures we remain in our reproducible era, which is better than the nemesis ever did! We specified a sample size of 100.
```{r}
#The Setup Block
## RUN ONCE: install.packages("devtools")
## RUN ONCE: devtools::install_github("sarahlotspeich/imputeCensRd", ref = "main")
library(imputeCensRd)
set.seed(114) 
censoring = "light" # censoring rate, other options included "moderate" or "heavy"
n = 100 # sample size, other options included 500, 1000, or 2000
```

This block generates the data for the experiment. It creates an uncensored covariate $Z$~Binomial(100,0.5), a covariate $X$~Weibull(0.75,0.25) that will eventually be censored, a random error vector $e$ simulated from a standard normal distribution, and specifies a true linear relationship and a response variable. $Q$ sets up a censoring condition to be used as a scale parameter in the censoring mechanism. When C > X, we consider an observation censored, as we have generated more than we can use. Changing the scale of the generating distribution (needs to be a distribution because censoring is random) changes the amount of overlap between W and C. $W$ compares $X$ and $C$ and outputs their minimums, and $D$ is an indicator that returns 1 if $X \leq C$. Since $D$ is an indicator, its mean is the percentage that isn't censored.
```{r}
#The Data Generation and ProofReading Block
z = rbinom(n = n, size = 1, prob = 0.5) # Uncensored covariate
x = rweibull(n = n, shape = 0.75, scale = 0.25) # To-be-censored covariate
e = rnorm(n = n, mean = 0, sd = 1) # Random errors
y = 1 + 0.5 * x + 0.25 * z + e # Continuous outcome
q = ifelse(censoring == "light", 2, 
           ifelse(censoring == "moderate", 0.35, 0.05)) # Rate parameter for censoring
c = rweibull(n = n, shape = 1, scale = q) # Random censoring mechanism
w <- pmin(x, c) # Observed covariate value
d <- as.numeric(x <= c) # "Event" indicator
dat = data.frame(z, w, y, d) # Construct dataset

# Check % censored
1 - mean(dat$d) # 12% 
```

This block writes a function to specify the true survival function (modeled by a Weibull distribution). A survival object in R needs Z as a parameter, because the function is generally S(X|Z). In this particular scenario, X and Z are independent. The lower tail parameter specifies we're looking at the P(X > x). Reminder, X is continuous.

```{r}
# Write a function for the true survival function used to generate Weibull X 
trueSURV = function(q, z) {
  pweibull(q = q, shape = 0.75, scale = 0.25, lower.tail = FALSE)
}
```

This block is copied from the imputeCensRd package on Github, because I couldn't figure out how the cmi_custom function worked without doing a little detective work. Someone call Sherlock, I'm coming for him! I've learned that Delta=0 indicates a censored observation, and we use similar variable naming conventions. The commented section on top is helpful to understand the function structure. If a $Z$ isn't specified, we assume the passed-in survival function has a default null $Z$. Uncens is a copy of Delta, and t-diff is the size of the subinterval (base of the trapezoid). 

```{r}
#' Custom conditional mean imputation (CMI) for a censored predictor with user-specified survival function
#'
#' Custom conditional mean imputation (CMI) for a censored predictor using (externally calculated) user-specified conditional survival.
#'
#' @param W Column name of observed predictor values (including censored opens).
#' @param Delta Column name of censoring indicators. Note that \code{Delta = 0} is interpreted as a censored observation.
#' @param Z Column name of additional fully observed covariates.
#' @param data Dataframe or named matrix containing columns \code{W}, \code{Delta}, and \code{Z}.
#' @param useSURV Assumed survival function for \code{W} given \code{Z}. The only arguments to \code{useSURV} should be \code{W} and \code{Z}, in that order.
#' @param trapezoidal_rule A logical input for whether the trapezoidal rule should be used to approximate the integral in the imputed values. Default is \code{FALSE}.
#'
#' @return
#' \item{imputed_data}{A copy of \code{data} with added column \code{imp} containing the imputed values.}
#' \item{code}{Indicator of algorithm status (\code{TRUE} or \code{FALSE}).}
#'
#' @export

cmi_custom <- function(W, Delta, Z, data, useSURV, trapezoidal_rule = FALSE) {
  # Calculate survival with original model coefficients using custom function
  if(is.null(Z)) {
    data <- data.frame(data, surv = useSURV(data[, W]))
  } else {
    data <- data.frame(data, surv = useSURV(data[, W], data[, Z]))
  }

  # Order data by W
  data <- data[order(data[, W]), ]

  # Create an indicator variable for being uncensored
  uncens <- data[, Delta] == 1

  # Calculate imputed values
  data$imp <- data[, W]
  if (trapezoidal_rule) {
    # Distinct rows (in case of non-unique obs values)
    data_dist <- unique(data[, c(W, Delta, Z, "surv")])

    # [T_{(i+1)} - T_{(i)}]
    t_diff <- data_dist[- 1, W] - data_dist[- nrow(data_dist), W]

    # Censored subject values (to impute)
    t_cens <- data[data[, Delta] == 0, W]

    # Follow formula assuming Accelerated Failure Time model for S(X|Z)
    for (x in which(!uncens)) {
      Cj <- data[x, W]
      Sj <- data_dist[-1, "surv"] + data_dist[- nrow(data_dist), "surv"]
      num <- sum((data_dist[-nrow(data_dist), W] >= Cj) * Sj * t_diff)
      denom <- data[x, "surv"]
      data$imp[x] <- (1 / 2) * (num / denom) + Cj
    }
  } else {
    ## Use integrate() to approximate integral from W to \infty of S(t|Z) (this is adaptive quadrature)
    if (is.null(Z)) {
      int_surv <- sapply(
        X = which(!uncens),
        FUN = function(i) {
          tryCatch(expr = integrate(f = function(t) useSURV(t), lower = data[i, W], upper = Inf)$value,
                   error = function(e) return(NA))
        }
      )
    } else {
      int_surv <- sapply(
        X = which(!uncens),
        FUN = function(i) {
          tryCatch(expr = integrate(f = function(t) useSURV(t, data[i, Z]), lower = data[i, W], upper = Inf)$value,
                   error = function(e) return(NA))
        }
      )
    }
    ## Calculate E(X|X>W,Z) = int_surv / surv(W|Z) + W
    data$imp[which(!uncens)] <- data[which(!uncens), W] + int_surv / data[which(!uncens), "surv"]
  }

  ## Check for infinite/NA imputed values
  if (any(is.na(data$imp))) {
    data$imp[which(is.na(data$imp))] <- data[which(is.na(data$imp)), W]
  }
  if (any(data$imp  == Inf)) {
    data$imp[which(data$imp == Inf)] <- data[which(data$imp == Inf), W]
  }

  # Return input dataset with appended column imp containing imputed values
  if (any(is.na(data$imp))) {
    return(list(imputed_data = data, code = FALSE))
  } else {
    return(list(imputed_data = data, code = TRUE))
  }
}
```

This block executes the imputation. It calculates conditional means with trapezoidal rule and adaptive quadrature, and this is done just by changing the rule parameter in the cmi_custom function from the imputeCensRd package.

```{r}
# Imputation approach 1: Calculate conditional means with trapezoidal rule
trap_rule_imp = cmi_custom(W = "w", Delta = "d", Z = "z", data = dat, useSURV = trueSURV, trapezoidal_rule = TRUE)
trap_rule_imp$code # If TRUE, imputation was successful 
trap_rule_fit <- lm(y ~ imp + z, data = trap_rule_imp$imputed_data) # Fit analysis model to imputed data

# Imputation approach 2: Calculate conditional means with adaptive quadrature
adapt_quad_imp = cmi_custom(W = "w", Delta = "d", Z = "z", data = dat, useSURV = trueSURV, trapezoidal_rule = FALSE)
adapt_quad_imp$code # If TRUE, imputation was successful 
adapt_quad_fit <- lm(y ~ imp + z, data = adapt_quad_imp$imputed_data) # Fit analysis model to imputed data

# Compare models 
trap_rule_fit
adapt_quad_fit

```



